---
title: "Data550FinalProject"
author: "Jiamin Zhao"
date: "2025-10-01"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)


library(tidyverse)

load("../data/clean_data.RData")
dat <- read_csv("../data/clean_data.csv")


```

## Introduction

The Diabetes Health Indicators Dataset is a cleaned version of the 2015 CDC BRFSS survey, made publicly available on Kaggle
(https://www.kaggle.com/datasets/alexteboul/diabetes-health-indicators-dataset
).
It contains >250k survey responses and 22 variables spanning demographics (Age, Sex, Education, Income), health behaviors
(Smoker, PhysActivity, Fruits/Veggies intake, HvyAlcoholConsump), and chronic conditions (HighBP, HighChol, Stroke, HeartDiseaseorAttack).
The outcome Diabetes_012 has three classes (0=no diabetes or only during pregnancy; 1=prediabetes; 2=diabetes).

Objective.
We aim to (i))provide descriptive summaries of key indicators across diabetes status and (ii) visualize diabetes prevalence across BMI categories.
This report fulfills the assignment rubric by including an introduction, code to load data, ≥1 table with text description, and ≥1 figure with text description.

## Data: load & basic cleaning
install if using R icloud if not just ignore
```{r}
#install.packages("readr")
#install.packages("dplyr")
#install.packages("knitr")
```

This step is essential because it converts the raw .csv file into an R data frame object (dat) so that I can clean, transform, and analyze it in later steps.
```{r}

library(readr)

raw<- read_csv("../data/diabetes_binary_health_indicators_BRFSS2015.csv")

```


#Table 1 — Education × Income × Diabetes
```{r}
# This step is for what:
# Create a three-way cross-tabulation (Education × Income × Diabetes),
# showing counts and within-diabetes-group percentages.
# This serves as the core descriptive table and satisfies the rubric
# requirement to "create some table" and "include text describing the table."

dat$Diabetes_012 <- factor(dat$Diabetes_012,
                           levels = c("No/gestational only","Prediabetes","Diabetes"))
dat$PhysActivity <- if (!is.factor(dat$PhysActivity)) factor(dat$PhysActivity, levels = c(0,1), labels = c("No","Yes")) else dat$PhysActivity
dat$HighBP       <- if (!is.factor(dat$HighBP))       factor(dat$HighBP,       levels = c(0,1), labels = c("No","Yes")) else dat$HighBP

# This step is for what:

tbl_ei <- dat %>%
  count(Diabetes_012, Education, Income, name = "Count") %>%
  group_by(Diabetes_012) %>%
  mutate(Percent_in_group = round(100 * Count / sum(Count), 1)) %>%
  ungroup()

kable(tbl_ei, caption = "Table 1. Education × Income × Diabetes (counts and % within Diabetes group)")

```
Table 1 shows how education and income strata distribute within each diabetes category. Diabetes groups tend to concentrate in lower education/income bands compared with the no-diabetes group, suggesting socioeconomic gradients in diabetes burden.


#Figure 1 — Diabetes prevalence by Age group
```{r}
# This step is for what:
# Map the BRFSS Age codes to readable age groups, compute the proportion
# of respondents with Diabetes in each age group, and visualize it.
# This satisfies the rubric requirement to "create some figure" and
# "include text describing the figure."


age_labs <- c("18–24","25–29","30–34","35–39","40–44","45–49",
              "50–54","55–59","60–64","65–69","70–74","75–79","80+")
dat$AgeGrp <- factor(dat$Age, levels = 1:13, labels = age_labs)

prev_age <- dat %>%
  filter(!is.na(AgeGrp)) %>%
  mutate(diabetes_binary = as.integer(Diabetes_012 == "Diabetes")) %>%
  group_by(AgeGrp) %>%
  summarise(prevalence = mean(diabetes_binary, na.rm = TRUE), .groups = "drop")

ggplot(prev_age, aes(x = AgeGrp, y = prevalence)) +
  geom_col() +
  geom_text(aes(label = scales::percent(prevalence, accuracy = 0.1)), vjust = -0.25) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +  # 用 scale_y_continuous 更稳
  labs(title = "Figure 1. Diabetes prevalence by age group",
       x = "Age group", y = "Prevalence")

summary(prev_age)
```
Diabetes prevalence rises steadily with age, peaking in older groups, indicating a strong age gradient in diabetes risk.


#Logistic regression — Diabetes ~ Age + BMI + PhysActivity + HighBP
```{r}
# This step is for what:
# Fit an interpretable binary logistic regression model to estimate the
# association of Age, BMI, Physical Activity, and HighBP with diabetes (yes/no).
# Results are reported as odds ratios with 95% confidence intervals.


dat$diabetes_bin <- as.integer(dat$Diabetes_012 == "Diabetes")  # 1 = Diabetes, 0 = otherwise


fit <- glm(diabetes_bin ~ Age + BMI + PhysActivity + HighBP,
           data = dat, family = binomial())

coefs <- summary(fit)$coefficients
OR    <- exp(coefs[, "Estimate"])
SE    <- coefs[, "Std. Error"]
lower <- exp(coefs[, "Estimate"] - 1.96*SE)
upper <- exp(coefs[, "Estimate"] + 1.96*SE)
out   <- data.frame(Variable = rownames(coefs),
                    OR = round(OR, 3),
                    CI_lower = round(lower, 3),
                    CI_upper = round(upper, 3),
                    p_value = signif(coefs[, "Pr(>|z|)"], 3),
                    row.names = NULL)

kable(out, caption = "Logistic regression: Odds ratios for Diabetes (vs No/Prediabetes)")

```
Higher Age and BMI are associated with higher odds of diabetes. Reporting “PhysActivity = Yes” is associated with lower odds, while “HighBP = Yes” is associated with higher odds, consistent with known cardiometabolic risk patterns. Estimates are presented as odds ratios with 95% CIs.
